Strict Pure Functional Programming and Automatic Parallelization and Transparent Fault Tolerance\\ (New project paper)

### 2016-03-29 pubs/tfp16/paper.tex
some related work

In the imperative world side-effect-free tasks could be safely restarted after
failure, needing only their well-defined inputs and not arbitrary access to
mutable global state.  The Legion system is being augmented with this
capability, and in such a way that it will be largely transparent to the programmer.
Ericsson's quasi-functional language Erlang and its runtime system was
designed for fault tolerance.~\cite{Cesarini:2009}.  It has long been
recognized that the pure parts of a functional program (excluding, e.g., the
I/O monad in Haskell) could be safely restarted.  The recent work of
[Trinder's student] impressively demonstrates this for an augmented ghc/runtime
on a large-scale distributed-memory system.

@@

Our project is the light-weight implementation of a pure, higher-order,
polymorphic, functional language and runtime system with which we can
experiment with automatic parallelization strategies with degrees of language
strictness ranging from lazy to fully strict, and secondarily, with mechanisms
for transparent fault tolerance.  Light-weight has several meanings or
implications.  First and foremost it means that \emph{we are not attempting to
  compete with ghc in any respect}---succumbing to the temptation of
recreating ghc language, type system, and runtime features, or significant
code optimization, would simply divert us from our central goals.\footnote{We
  estimate that we will have, to within an order of magnitude, one person-hour
  of total available development time for each person-year that ghc currently
  represents, or put another way, one second for each hour, respectively.}

@@

ghc 8.0 gives three possibilities: conventional non-strict, constructor
strict, and function- and constructor strict, but there are possible
variations.  The obvious variation is function, but not constructor,
strictness.  In Haskell the distinction between constructors and functions is
somewhat blurred because constructors behave like functions (outside of
pattern matching), but this need not be the case.

@@

Our wish list sounds much like the ghc Haskell implementation, and indeed ghc
is part of our master plan.  In short, our plan is to use ghc as our front-end
to generate its intermediate form, STG~\cite{PJ-stockhardware}, then escape to
our system from there.  There are various possible ways to accomplish the
former: using ghc's option to dump STG as text, hacking ghc itself and,
optionally, using ghc as a library.

@@

Our incarnation of STG, as is our presentation in Figure~\ref{STGsyntax} which
is meant to evoke both an abstract and concrete syntax, are very similar to
that given in Marlow and Peyton Jones' \emph{fast curry}
paper~\cite{Marlow:2004}, to which we refer so many times that we will
subsequently refer to it as simply \emph{EA}.\footnote{\emph{Eval/Apply} from
  the title because \emph{FC} likely has another connotation to the reader.}

@@

\end{tabular}
\end{centering}
\caption{STG syntax}
\label{STGsyntax}
\end{figure}

@@

In STG---one of ghc's
intermediate representations---the distinction is clear: heap objects are
explicitly constructed by \texttt{let} constructs in which constructor names
are given explicitly (not aliased); constructor names are not functions.

@@

\subsubsection{Proper tail calls.}
While performance is not a high priority, we attain proper tail calling instead of
using the trampoline used in earlier versions of ghc~\cite{PJ-stockhardware}.
Because we maintain our own data and control (continuation) stack all
generated C functions take no explicit arguments and have return type void,
and with a single exception in the non-strict case are invoked in tail call
positions, at non-zero optimization levels both gcc and Clang/LLVM generate
jumps instead of calls for C calls in tail call positions.  When using gcc
we also use a fixed set of machine registers for passing/returning values.

@@

\subsection{Mini-Haskell}
As mentioned, a current near-term goal is to use ghc as our front-end to
generate STG as input to our system.  In the meantime, as a stop-gap measure
to make the generation of test cases easier, we have implemented a
\emph{mini-Haskell} front-end that supports a subset of standard Haskell
syntax with \emph{MagicHash} enabled, with a single extension.  Recall that
\emph{MagicHash} allows \texttt{\#} to be a suffix of an identifier, or denote an
unboxed literal value.  Our extension (to STG as well)
is the provision of the optional \texttt{unboxed} keyword in datatype definitions
to allow user-defined unboxed data types, e.g.,
\begin{verbatim}
  data unboxed Bool# = False# | True# .
\end{verbatim}

@@

An informal specification of the mini-Haskell grammar is given in
Figure~\ref{miniHaskell}.  There is significant overlap with STG resulting
from the structural similarities, but the redundancies are included for
completeness.

@@

\end{tabular}
\label{miniHaskell}
\caption{mini-Haskell syntax}
\end{figure}

@@

\subsubsection{The Layout Rule}
Mini-Haskell implements Haskell's Layout Rule in an almost direct translation of
the specification given in section 10.3 of the Haskell 2010 Report.  The notable
exception is the omission of any attempt to define the \texttt{parse-error}
function.  This is apparently a tricky spot for Haskell compilers (even GHC has
not always had it right) so, in the name of tradition, it remains a tricky spot
for our mini-Haskell front end.  The edge cases that it will incorrectly handle
as a result of this are exactly that: edge cases.  As such, they are easy to
avoid in practical usage.

The inclusion of the layout rule may seem small and insignificant, but it
allows its users to produce code that is visually similar to Haskell code.
This makes it easier to read and write for any Haskell programmer who relies
on layout to delineate context, presumably the vast majority.  It also allows
for the use of Haskell programming environments that support ``smart''
indentation, e.g., \emph{haskell-mode} for Emacs.

@@

No bridge from ghc STG to our STG yet.

At the time of writing we have a complete serial implementation, including
\begin{itemize}
\item STG (inc. ADT w/unboxed) parser
\item type inferencer
\item code generator
\item runtime system including a garbage collector
\end{itemize}

We have implemented a simple copying garbage collector \cite{Cheney:1970} with
small \emph{nurseries} in the multithreaded case \cite{Marlow:2008}. We have
not focused on the development of a more sophisticated garbage collector
because the goal of this project is to explore automatic parallelization.

@@

We perform standard Hindley-Milner type inference and enforcement, augmented
with built-in and user-defined unboxed types as described by Peyton-Jones and
Launchbury's seminal work~\cite{Jones91unboxedvalues}, but for unboxed types code
generation is currently limited to built-in single-word (64-bit) or smaller
types (e.g., \texttt{Int\#}, \texttt{UInt\#}, \texttt{Float\#}, \texttt{Double\#})
and user-defined enumerations, e.g., \texttt{data unboxed Bool\# = False\# |
  True\#}.

@@

We have developed a extensive test suite which uses the CMake \cite{cmake-book}
CTest tool. The full test suite is automatically run, and any failures reported,
after every comment to the version control system. We use a Jenkins continuous
integration server \cite{jenkins-book} to automate this build and testing process.
The (git) version control repository is hosted externally so our student
collaborators can contribute at their leisure.

There are several hundred small STG and mini-Haskell programs in the test suite.
It is partitioned into a number of directories which control what
testing is done for a specific program. Given our goal of exploring various
levels of strictness, most tests are run with all evaluation strategies (3
degrees of function strictness by two degrees of constructor strictness).
One interesting class of test programs are those which require non-strict
semantics. For these programs we test not only that the test leads to the expected
result with non-strict semantics, but also that the program returns a
"blackhole" in the case of strict semantics. Students who were Haskell/STG
beginners have contributed substantially to the test suite. We have also found
that the test suite has be of use as a learning tool for these students to
discover  what level of non-strictness was required for their programs,
a benefit that we did not initially envision. Over time we have developed
extensive STG and mini-Haskell preludes of common programming patterns, to
aid in the writing of test programs.

@@

We also run each test with various levels of garbage collection enabled, this
has helped us to find bugs not only in the garbage collector, but also in the
code generator.

The various enumerations of strictness and garbage collection levels mean that
we run a quite large number of tests. However, this is fully automated via the
testing and continuous integration infrastructure.

@@

\section{Closely Related Work}

Clearly various papers describing some of the inner workings of ghc at various point of its
evolution have greatly informed our approach and these have all be previously cited.

In his BSc thesis, Don Stewart described an implementation wherein ghc was used as a Haskell
compiler front-end, dumping STG code as text, parsing that text, then translating to code for
execution on the Java Virtual Machine.~\cite{Stewart-BSc}.

Nvidia research compiler.

\section{Summary, Future Work, Conclusions?}

@@

Funding for this project and its participants has been provided by the DOE
NNSA LANL Laboratory Directed Research and Development program award
20150845ER; the National Science Foundation Science, Technology, Engineering,
and Mathematics Talent Expansion Program (STEP) program for undergraduate
students; and the DOE Science Undergraduate Laboratory Internship (SULI) program.
%
Los Alamos National Laboratory is managed and operated by Los Alamos National
Security, LLC (LANS), under contract number DE-AC52-06NA25396 for the
Department of Energy’s National Nuclear Security Administration (NNSA).

% just get these into the bib
\cite{ghc-as-library}
\cite{Heren02}

>>>>

In the imperative world side-effect-free tasks could be safely restarted after
failure, needing only their well-defined inputs and not arbitrary access to
mutable global state.  The Legion system is being augmented with this
capability, and in such a way that it will be largely transparent to the programmer.
Ericsson's quasi-functional language Erlang and its runtime system was
designed for fault tolerance.~\cite{Cesarini:2009}.  It has long been
recognized that the pure parts of a functional program (excluding, e.g., the
I/O monad in Haskell) could be safely restarted.  The recent work of
[Trinder's student] impressively demonstrates this for an augmented GHC/runtime
on a large-scale distributed-memory system.

@@

Our project is the light-weight implementation of a pure, higher-order,
polymorphic, functional language and runtime system with which we can
experiment with automatic parallelization strategies with degrees of language
strictness ranging from lazy to fully strict, and secondarily, with mechanisms
for transparent fault tolerance.  Light-weight has several meanings or
implications.  First and foremost it means that \emph{we are not attempting to
  compete with GHC in any respect}---succumbing to the temptation of
recreating GHC language, type system, and runtime features, or significant
code optimization, would simply divert us from our central goals.\footnote{We
  estimate that we will have, to within an order of magnitude, one person-hour
  of total available development time for each person-year that GHC currently
  represents, or put another way, one second for each hour, respectively.}

@@

GHC 8.0 gives three possibilities: conventional non-strict, constructor
strict, and function- and constructor strict, but there are possible
variations.  The obvious variation is function, but not constructor,
strictness.  In Haskell the distinction between constructors and functions is
somewhat blurred because constructors behave like functions (outside of
pattern matching), but this need not be the case.

@@

Our wish list has much in common with the GHC Haskell implementation, and
indeed GHC is part of our master plan.  In brief, the phases of the GHC
compiler are 1) parsing and type checking; 2) transformation to the GHC
\emph{Core} intermediate representation; 3) analysis and transformation on
Core; 4) transformation of Core to the \emph{STG} intermediate representation;
5) transformation of STG to machine code via one of various imperative
representations (Cmm, LLVM, C).

In short, our plan is to use GHC as our front-end to generate its intermediate
form, STG~\cite{PJ-stockhardware}, then escape to our system from there.
There are various possible ways to accomplish the former: using GHC's option
to dump STG as text, hacking GHC itself and, optionally, using GHC as a
library~\cite{ghc-as-library}.  An alternative would be to start with GHC Core.

@@

Our incarnation of STG, as is our presentation in Figure~\ref{fig:STGsyntax} which
is meant to evoke both an abstract and concrete syntax, are very similar to
that given in Marlow and Peyton Jones' \emph{fast curry}
paper~\cite{Marlow:2004}, to which we refer so many times that we will
subsequently refer to it as simply \emph{EA}.\footnote{\emph{Eval/Apply} from
  the title because \emph{FC} likely has another connotation to the reader.}

@@

\end{tabular}
\end{centering}
\caption{STG syntax}
\label{fig:STGsyntax}
\end{figure}

@@

In STG---one of GHC's
intermediate representations---the distinction is clear: heap objects are
explicitly constructed by \texttt{let} constructs in which constructor names
are given explicitly (not aliased); constructor names are not functions.

@@

\subsubsection{Proper tail calls.}
While performance is not a high priority, we attain proper tail calling instead of
using the trampoline used in earlier versions of GHC~\cite{PJ-stockhardware}.
Because we maintain our own data and control (continuation) stack all
generated C functions take no explicit arguments and have return type void,
and with a single exception in the non-strict case are invoked in tail call
positions, at non-zero optimization levels both gcc and Clang/LLVM generate
jumps instead of calls for C calls in tail call positions.  When using gcc
we also use a fixed set of machine registers for passing/returning values.

@@

\subsection{Mini-Haskell}
As mentioned, a current near-term goal is to use GHC as our front-end to
generate STG as input to our system.  In the meantime, as a stop-gap measure
to make the generation of test cases easier, we have implemented a
\emph{mini-Haskell} front-end that supports a subset of standard Haskell
syntax with \emph{MagicHash} enabled, with a single extension.  Recall that
\emph{MagicHash} allows \texttt{\#} to be a suffix of an identifier, or denote an
unboxed literal value.  Our extension (to STG as well)
is the provision of the optional \texttt{unboxed} keyword in datatype definitions
to allow user-defined unboxed data types, e.g.,
\begin{verbatim}
  data unboxed Bool# = False# | True# .
\end{verbatim}

@@

An informal specification of the mini-Haskell grammar is given in
Figure~\ref{fig:miniHaskell}.  There is significant overlap with STG resulting
from the structural similarities, but the redundancies are included for
completeness.

@@

\end{tabular}
\label{fig:miniHaskell}
\caption{mini-Haskell syntax}
\end{figure}

@@

\subsubsection{The Layout Rule}
Mini-Haskell implements Haskell's layout rule in an almost direct translation
of the specification given in section 10.3 of the Haskell 2010
Report.~\cite{haskell2010report} The notable exception is the omission of any
attempt to define the \texttt{parse-error} function.  This is apparently a
tricky spot for Haskell compilers (even GHC has not always had it right) so,
in the name of tradition, it remains a tricky spot for our mini-Haskell front
end.  The edge cases that it will incorrectly handle as a result of this are
exactly that: edge cases.  As such, they are easy to avoid in practical usage.

The inclusion of the layout rule may seem small and insignificant, but it
allows users to produce code that is visually similar to Haskell code.
This makes it easier to read and write for any Haskell programmer who relies
on layout to delineate context, presumably the vast majority.  It also allows
for the use of Haskell programming environments that support ``smart''
indentation, e.g., \emph{haskell-mode} for Emacs.~\cite{haskellmode}.

@@

No bridge from GHC STG to our STG yet.

At the time of writing we have a complete serial implementation, including
\begin{itemize}
\item STG (inc. ADT w/unboxed) parser
\item type inferencer
\item various analysis and transformation passes
\item code generator
\item runtime system including a garbage collector
\end{itemize}

We have implemented a simple copying garbage collector~\cite{Cheney:1970} with
small \emph{nurseries} in the multithreaded case \cite{Marlow:2008}. We have
not focused on the development of a more sophisticated garbage collector
because the goal of this project is to explore automatic parallelization.

@@

We perform standard Hindley-Milner type inference and
enforcement~\cite{Heren02}, augmented with built-in and user-defined unboxed
types as described by Peyton-Jones and Launchbury~\cite{Jones91unboxedvalues},
but for unboxed types code generation and runtime support is currently limited
to built-in single-word (64-bit) or smaller types (e.g., \texttt{Int\#},
\texttt{UInt\#}, \texttt{Float\#}, \texttt{Double\#}) and user-defined
enumerations, e.g., \texttt{data unboxed Bool\# = False\# | True\#}.

@@

We have developed a extensive test suite which uses the CMake CTest
tool.~\cite{cmake-book} The full test suite is automatically run, and any
failures reported, after every commit to the Git revision control system. We
use a Jenkins continuous integration server to automate this build and testing
process.~\cite{jenkins-book} The Git repository is hosted externally so our
student collaborators can contribute at their leisure.

There are several hundred small STG and mini-Haskell programs in the test suite.
It is partitioned into a number of directories which control what
testing is done for a specific program. Given our goal of exploring various
levels of strictness, most tests are run with all evaluation strategies (three
degrees of function strictness by two degrees of constructor strictness).
One interesting class of test programs are those which require lazy
semantics. For these programs we test not only that the test leads to the expected
result with non-strict semantics, but also that the program returns a
\emph{BLACKHOLE} in the case of strict semantics. Students who were Haskell/STG
beginners have contributed substantially to the test suite. We have also found
that the test suite has be of use as a learning tool for these students to
discover what level of non-strictness was required for their programs,
a benefit that we did not initially envision. Over time we have developed
extensive STG and mini-Haskell preludes of common programming patterns to
aid in the writing of test programs.

@@

We also run each test with various levels of garbage collection enabled. This
has helped us to find bugs not only in the garbage collector, but also in the
code generator.  Interestingly, in practice garbage collection can both mask
and expose bugs.

The various enumerations of strictness and garbage collection levels mean that
we run a quite large number of tests (currently order thousands). However,
this is fully automated via the testing and continuous integration
infrastructure.

@@

\section{Related Work}

Various papers describing some of the inner workings of GHC at various point
of its evolution have greatly informed our approach and these have all been
previously cited.

\subsubsection{\emph{Strict but still pure} is not new.}
Very simple implementations have been prototyped~\cite{explicit-lazy}.  
%
A plug-in for GHC (prior to 8.0) exists for making functions, but not data
structures, strict.~\cite{strict-ghc-plugin}.
%
The Disciplined Disciple Compiler is a full-blown implementation that has been
underway for some time, but the emphasis appears to be on a much more complex
type system for handling stateful computation, and the prospects for its
eventual completion are unknown~\cite{disciple}.
%
NVIDIA's NOVA language, and corresponding compiler, implements a strict
(``call by value''), pure, polymorphic, higher-order functional language with
built-in parallel operators (e.g., \emph{map}, \emph{reduce}, and
\emph{scan}).~\cite{NVIDIA:2013} Three back-ends were implemented: for
sequential C, parallel (multi-threaded) C, and CUDA C, the lattermost
targeting NVIDIA GPUs.

\subsubsection{Non-strictness as antithetical to parallel evaluation.}  While
compile-time analysis can conservatively estimate when it is safe to strictify
and parallelize, in general the problem is undecidable and the compiler must
default to purely data-dependency-driven evaluation on demand.  This has been
shown to be problematic for three reasons~\cite{Mattson94}.  First, it is
often desired to force parallel evaluation beyond simply to WHNF, requiring
elaborate strictification as previously discussed.  Second is a space
management and runtime issue: because non-strict evaluation is generally
space-intensive, the runtime may be forced to prematurely abort parallel
evaluation when garbage collection is triggered by space constraints.  Third,
when aborting a speculative evaluation, to be able to reclaim space sufficient
bookkeeping needs to be done to be able to roll back the reduction graph.  The
conclusion was that it is hard to amortize the runtime overhead over gains in
parallelism, and thus justify either the runtime or application implementation
efforts in the non-strict context.  In the strict context the runtime can
simply parallelize until resources are saturated, and much less bookkeeping is
required to recover space when speculative evaluations are to be terminated.

\subsubsection{GHC as a front-end.}
The Intel Labs Haskell Research Compiler (HRC) is the most closely related
work to ours of which we are aware.~\cite{Liu:2013,Petersen:2013} As in our
(planned) approach they use GHC as the front end, but exit the GHC toolchain
at the final GHC Core stage, just before transformation to STG.  The essence
of their effort is to target an existing, highly optimizing, Intel functional
language compiler that is ``largely language agnostic'' and performs (among
many other optimization) SIMD vectorization.

In his BSc thesis, Don Stewart described an implementation wherein GHC was
used as a Haskell compiler front-end, dumping STG code as text, parsing that
text, then translating to code for execution on the Java Virtual
Machine.~\cite{Stewart-BSc}.

\subsubsection{Fault tolerance for functional programming.}
Pointon et al.\ developed a limited fault tolerance mechanism for Glasgow
distributed Haskell based on distributed exception
handling.~\cite{Pointon-GdH-00}.  However, it appears that their scheme for
handling failed processing elements correctly~\cite{} may have never been
implemented.

Recently, Robert Stewart extended Haskell Distributed Parallel Haskell
(HdpH)~\cite{hdph} to be fault resilient at an impressive scale---1400 cores
on a distributed memory system.  Here the key recovery strategy is task
replication.~\cite{Stewart:2013}.

Trinder--speculative parallelism

Trinder's paper (or just web site?) on fault tolerance.

Trinder's student's thesis on fault tolerance.

\section{Future Work}

An almost unbounded amount of interesting work suggests itself; one such path
would lead to the recreation of GHC.  However, given the miniscule budget (and
thus the emphasis on student labor) we must maintain sharp focus on the main
goals:  implementing and evaluating mechanisms for automatic parallelism and
fault tolerance.

Currently we are developing the parallel runtime.  This summer (2016) a
returning student will (partially) bridge the gap between GHC STG and our STG,
and augment our runtime as needed in this effort.  Just as in the Intel Haskell
Research Compiler effort, we will not

@@

Funding for this project has been provided by the DOE NNSA LANL Laboratory
Directed Research and Development program award 20150845ER; the National
Science Foundation Science, Technology, Engineering, and Mathematics Talent
Expansion Program (STEP) program for undergraduate students; and the DOE
Science Undergraduate Laboratory Internship (SULI) program.
%
Los Alamos National Laboratory is managed and operated by Los Alamos National
Security, LLC (LANS), under contract number DE-AC52-06NA25396 for the
Department of Energy’s National Nuclear Security Administration (NNSA).
